# Introduction {#intro}

## What is statistical modeling and what do we need this for?

Typically, one simplifies the complex reality (and loses information) in order to make it better
understandable, mathematically treatable and to make predictions. 

Underlying our models, there are theories which should be [falsifiable](https://en.wikipedia.org/wiki/Falsifiability)
and testable.
For instance, I would be really surprised if I pull up my multimeter and measure the voltage (V) and
electric current (I) at a resistence (R) in a circuit and find that [Ohm's law](https://en.wikipedia.org/wiki/Ohm%27s_law) $V = IR$ is not true. 
This [**law**](https://en.wikipedia.org/wiki/Scientific_law)
can be tested over and over again and if one would find a single valid counterexample, 
the law would be falsified. It is also true that the law is probably not 100% accularate,
but an extremely good approximation of reality. Real-world measurements carry 
measurement errors and when plotting the data, one would see that the data points
might not lie exactly on a straight line. This is not a problem.

A [statistical model](https://en.wikipedia.org/wiki/Statistical_model) 
is a mathematical framework that represents the 
relationships between variables, helping us understand, infer, and
predict patterns in data. It acts as a bridge between observed data 
and the real-world processes that generated them. In health research, 
where variability and uncertainty are inherent, statistical models are 
valuable tools for making sense of complex phenomena. 
You can watch [this](https://www.youtube.com/watch?v=3d5ivs_8amQ&ab_channel=VeryNormal) as short intro.

Depending on the task at hand, we would use different models.
In any case, logical reasoning and critical thinking comes first, 
then comes the model. **It makes no sense to estimate statistical models just for the sake of it**.

**All models are wrong, but some are useful**. 
Or to quote [George Box](https://www.tandfonline.com/doi/abs/10.1080/01621459.1976.10480949):

> "Since all models are wrong the scientist cannot obtain
> a 'correct' one by excessive elaboration. On the contrary
> following William of Occam he should seek an economical
> description of natural phenomena. Just as the ability to
> devise simple but evocative models is the signature of the
> great scientist so overelaboration and overparameterization is often the mark of mediocrity."

In my opinion, statistical modeling is an art form: difficult and beautiful.

**One goal of this course** is to improve interpretation and limitations of statistical models.
They are not magical turning data into truth. Firstly, the rule gargabe in, garbage out (GABA) applies.
Secondly, statistical models are based on data and their variability and have inherent limitations
one cannot overcome even with the most sophisticated models. This is expressed for instance 
in the so-called [bias-variance trade-off](https://en.wikipedia.org/wiki/Bias%E2%80%93variance_tradeoff).
You can't have it all.

### Explanatory vs. Predictive Models

I can recommend reading [this](https://projecteuclid.org/journals/statistical-science/volume-25/issue-3/To-Explain-or-to-Predict/10.1214/10-STS330.full) 
article by Shmueli et al. (2010) on this topic.

Statistical models serve different purposes depending on the research question. Two primary goals are **explanation** 
and **prediction**, and each requires a different approach:

**Explanatory Models** focus on understanding causal relationships. 
These models aim to uncover mechanisms and answer **"why"** 
questions. For example:
   
   - Does smoking increase the risk of lung cancer? **Yes**. (If you want to see what a large effect-size looks like, check out [this study](https://bmjopen.bmj.com/content/bmjopen/8/10/e021611.full.pdf).)
   - How large is the "effect" of smoking on lung cancer? **Large**.
   - Does pain education and graded sensorimotor relearning improve disability (a question we ask in 
   our [Resolve Swiss project](https://data.snf.ch/grants/grant/220585))?

Explanatory models are **theory-driven**, designed to test hypotheses. Here, one wants to understand the underlying
mechanisms and the relationships between variables and hence often uses (parsimonious) models that are more interpretable, 
like linear regression.

**Predictive Models** prioritize forecasting future outcomes based on patterns in the data. 
These models aim to answer **"what will happen?"** For instance:
   
   - [Gait analysis](https://www.tandfonline.com/doi/abs/10.1080/03091902.2020.1822940) using Machine Learning (ML)?
   - [Skin cancer detection](https://jamanetwork.com/journals/jamadermatology/fullarticle/2756346) using neural networks?

Predictive models are **data-driven**, often using complex algorithms to achieve high accuracy. 
Their success is measured using metrics like [Root Means Square Error](https://computersciencewiki.org/index.php/Root-mean-square_error_(RMSE)) 
(RMSE), [Area Unter the Curve](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#:~:text=The%20area%20under%20the%20curve,ranks%20higher%20than%20'negative') 
(AUC), or **prediction error on new, unseen data**.
Any amount of model complexity is allowed. One could for instance estimate a 
[neural network](https://en.wikipedia.org/wiki/Neural_network_(machine_learning)) ("just" another statistical model) 
with many hidden layers and neurons in order to improve prediction quality. Interpretability of the model weights is not a priority here.

While explanatory and predictive goals often complement each other, 
their differences highlight the importance of clearly defining the purpose 
of your analysis. In applied health research, explanatory models help identify 
causal mechanisms, while predictive models can guide real-world decisions by 
providing actionable forecasts. Together, they enhance both our understanding 
of phenomena and our ability to make informed decisions in complex environments.

### Individual vs. Population Prediction

Another important distinction is between **individual vs. population** prediction.
In the smoking example above, we can be very sure about the mean effects that smoking has on lung cancer.
On an individual level, it is [harder to predict the outcome](https://www.liebertpub.com/doi/10.1089/rej.2019.2298). 
Nevertheless, individual predictions will be (notably) better than random guessing. We will discuss this in greater detail.

### Practical Use of Statistical Models

In my optinion, we should never be afraid to test our statistical models (as honestly as possible) against reality.
We could for instance ask ourselves: 

- "How much better does this model classify than the arithmetic mean? 
  (i.e., the linear model with just an intercept)"

- "How much better does this model classify than random guessing?"

- Is it worth the effort to collect data and estimate this model by using hundreds of hours of our time?

In some cases, these questions can be answered straightforwardly. 

- In advertising (Google, Facebook, ...), a couple of percentage points in prediction quality might make a difference of millions 
of dollars in revenue offsetting the statistitians salary. 

- Improved forecasts of a few percentage points in the stock market or just being [slightly better](https://en.wikipedia.org/wiki/Jim_Simons) 
than the average, will make you faboulously rich.

- Improved cancer forecasting might save lives, money and pain and is not only measured in money.

### Start at the beginning

What do we actually want to do in general? Very broadly speaking we want to:
**describe** the association of variables to each other that carry variability. 
Hence, the relationship is not deterministic like $$y = 2x + 3$$ but rather we need
to "loosen up" the relationship to account for variability (in $x$ and $y$).
So, the values $2$ and $3$ are not fixed but aflicted with uncertainty.
Depending on your philosophical view, you might say you want to find
the "true" but unknown relationship between variables.
This is what we do in simulation studies all the time: We know the true relationship, 
simulate data by adding variability and then try to estimate the true relationship we assumed in the first place.
For some practical applications, we can get a really nice and complete answer to our question
(for instance sample size for proportions).

So we are looking for a function $f$ such that

$$ Y = f(X) $$

where 

- $Y$ is the "outcome", "dependent variable" or "response". 
- $X$ are the "predictors". $X$ can be a single Variable $x$ or many 
variables $x_1, x_2, \ldots, x_p$.

It is important to be aware of the notation here:
"Predict" does **not necessarily** mean that we can predict the value in 
the future. It merely means we estimate the value (or mean) of $Y$ given $X$.

- This can be done at the same time points, known as **cross-sectional** analysis ("What is the maximum jumping height 
of a person given their age at a certain point in time, whereas both variables are measured at the same time?");
- or at different time points, known as **longitudinal analysis** ("What is the maximum jumping height of a person 10 years later ($t_2$)
 given their baseline health status at time $t_1$?").

The **simplest statistical model** would be the mean model where $Y$ is "predicted" by a 
constant: $Y = c$ which (at least in the classical linear regression) turns out to be $c = \bar{x}$. 
This simple model is often surprisingly good, or, to put it in other words, models with more complexity
are often not that much better with regards to multiple metrics.

## A (simple) model for adult body heights in the Bayesian framework

As repetition, read the parts about [Bayes statistics from QM1](https://jdegenfellner.github.io/Script_QM1_ZHAW/bayes_statistics.html) 
again to refresh your memory about the Bayesian framework.

It's recommendable to read the beginning of the book [Statistical rethinking](https://civil.colorado.edu/~balajir/CVEN6833/bayes-resources/RM-StatRethink-Bayes.pdf)
up until page 39 as well. We are not completely new to the topic of Bayes due 
to QM1.

We want to **start building our first model** right away.

Let's begin with the example in 
[Statistical rethinking](https://civil.colorado.edu/~balajir/CVEN6833/bayes-resources/RM-StatRethink-Bayes.pdf) 
using data from the [!Kung San](https://en.wikipedia.org/wiki/%C7%83Kung_people) 
people starting on page 79.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
library(rethinking)
data("Howell1")
d <- Howell1
str(d)
d2 <- d[d$age >= 18, ] # only adults
```
We want to model the adult height of the !Kun San people 
using prior knowledge (about the Swiss population) and data.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
library(tidyverse)
d2 %>% ggplot(aes(x = height)) + geom_histogram()
```

Since we already have domain knowledge in this area, we can say that heights are usually normally distributed,
or at least a mixture of normal distrubutions (female/male).
We assume the following model:
$$h_i \sim \text{Normal}(\mu, \sigma)$$

As in [QM1](https://jdegenfellner.github.io/Script_QM1_ZHAW/), 
we want to start with a Bayesian model and hence, we need some priors.

Since we are in Switzerland and just for fun, we use the mean of
[Swiss body heights](https://www.bfs.admin.ch/asset/de/30305714) 
as expected value for the **prior for the mean**. 
According to the link (Bundesamt für Statistik), 
the mean height of $n=21,873$ people in the Swiss sample is
$171.1$ cm. We choose the same $\sigma$ for the prior of the normal
as in the book not to deviate too much from the example at hand.

Next comes our **model definition in the Bayesian framework**, which
I often find more intuitive than the frequentist approach:

\[
h_i \sim \text{Normal}(\mu, \sigma)
\]
\[
\mu \sim \text{Normal}(171.1, 20)
\]
\[
\sigma \sim \text{Uniform}(0, 50)
\]

**Description of the model definition**: The heights are normally distributed with unknown mean and 
standard deviation. As our current knowledge about the mean height, we use 
a prior distribution for the mean (we do not know but want to estimate) by 
assuming the mean of a population we know and a standard deviation of $20$ cm which
allows are rather large range of possible values for $\mu$.
$\sigma$ is also unknown and a priori we restrict ourselves to values 
between $0$ and $50$ cm, whereas we assign equal plausibility to all 
values in this range (which can and should be critically discussed).

**Vizualisation of the model structure**:
```{r, echo=FALSE, , message=FALSE, warning=FALSE}
library(DiagrammeR)
grViz("
digraph BayesianModel {
  graph [layout = dot]
  
  h_i [label = 'h[i]\\n~ Normal(μ, σ)', shape = oval]
  μ [label = 'μ\\n~ Normal(171.1, 20)', shape = oval]
  σ [label = 'σ\\n~ Uniform(0, 50)', shape = oval]

  μ -> h_i
  σ -> h_i
}
")
```

Mind that there is a conceptual difference between the normal distribution of the heights 
and the normal prior distribution of the mean. The latter expresses our prior knowledge/insecurity
about the unobserved mean. The normal distribution says we expect the heights to be normally distributed
but we do not know the parameters ($\mu$ and $\sigma$) yet.

Of course we would not need the prior here due to the large sample size, 
but let's do it anyways for demonstration purposes.
We are not completely uninformed about body heights and express our 
knowledge with the prior for $\mu$.
The $20$ in the prior for the mean expresses our range of possible true 
mean values and aknowledge
that there are a variety of different subpopulations with different means.

Using the Swiss data in the link one could estimate that the standard deviation of the heights 
from $21,873$ Swiss people is around is $25.6553$ cm ([Exercise 1](#exercise1_Intro)).

Remember, in the Baysian world, there is no **fixed but unknown**
parameter, but instead we define a distribution over the unobserved parameter.

We **visualize the prior for $\mu$**.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
curve(dnorm(x, 171.1, 20), from = 100, to = 250)
```

The **prior for $\sigma$** is uniform between $0$ and $50$ cm. This is a very wide prior and
just constrains the values to be positive and below $50$ cm. 
This could be stronger of course.

**Visualization of the prior for $\sigma$**:

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
curve(dunif(x, 0, 50), from = -10, to = 60)
```

Note, we didn’t specify a prior probability distribution of heights
directly, but once we’ve chosen priors for $\mu$ and $\sigma$, these imply a 
prior distribution of individual heights. 

**Without** even having seen the **new data**, we can check what our prior
(model) for heights would predict. This is important. If the prior already 
predicts impossible values, we should reconsider our priors and/or model.

So, we simply draw $\mu$ and $\sigma$ from the priors and then draw heights 
from the normal distribution using the drawn parameters.

**Vizualisation of the prior for heights**:

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
sample_mu <- rnorm(10^4, 171.1, 20)
sample_sigma <- runif(10^4, 0, 50)
prior_h <- rnorm(10^4, sample_mu, sample_sigma)
length(prior_h)
dens(prior_h)
```

The prior is not itself a Gaussian distribution, but a distribution of
relative plausibilities of different heights, before seeing the data.

Now, there are a couple of different ways to estimtate the model incorporating 
the new data. For didactic reasons, grid approximation is often used (as in the book).
For many parameters, grid approximation becomes more and more infeasible (due to combinatorial explosion). 

We will skip that for now and use quadratic approximation instead which
works well for many common procedures in applied statistics (like linear regression).
Later, you'll probably use (or the software in the background) mostly Markov 
chain Monte Carlo (MCMC) sampling to get the posterior. 
[Pages 39 and the following](https://civil.colorado.edu/~balajir/CVEN6833/bayes-resources/RM-StatRethink-Bayes.pdf) 
explain the 3 concepts grid approximation, quadratic approximation and MCMC.

In short, **quadratic approximation** assumes that our posterior distribution 
of body heights can be approximated well by a normal distribution, 
at least near the peak. 

Please read the [addendum](#bivariate_normal) to get a clearer picture of 
what a bivariate normal distribution is.

Using the library `rethinking` we can estimate the model using quadratic approximation.
First, we define the model in the `rethinking` syntax (see R code 4.25 in the book).

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
library(rethinking)
flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(171.1, 20),
  sigma ~ dunif(0, 50)
)
```

Then we estimate/fit the model using quadratic approximation.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
m_heights <- quap(flist, data = d2)
```

Now let's take a look at the fitted model:
(Note: In the online-version of the book, they used the command `map` instead of `quap`.)

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
precis(m_heights)
```
Above, we see the mean of the posterior for $\mu$ **and** $\sigma$;
and a 89% credible interval for those parameters.

We can now plot the posterior distribution of the mean and the standard deviation.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
post <- extract.samples(m_heights, n = 10^4)
head(post)
dens(post$mu)
dens(post$sigma)
```

Note, that **these samples come from a multi-dimensional posterior distribution**.
In our case, we approximated the posterior with a 
[bivariate normal distribution](https://en.wikipedia.org/wiki/Multivariate_normal_distribution).
They are not necessarily independent from each other, but in this case they are.
We know this from the model definition above. $\mu$ and $\sigma$ are both
defined as normal respectively uniform distributions and by definition do not
influence each other. This is also visible in the vizualisation of the model structure: 
There is no confounding variable or connection between those priors. One could
think of a common variable $Z$ that influences both $\mu$ and $\sigma$. This could
be genetic similarity which could influence both $\mu$ and $\sigma$.

Let's verify that $\mu$ and $\sigma$ are uncorrelated:

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
vcov(m_heights)
```
gives you the variance-covariance matrix of the parameters of the posterior
distribution. In the diagonal you see the variance of the parameters.

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
diag(vcov(m_heights))
```
And we can compute the correlation matrix easily:

```{r, echo=TRUE, , message=FALSE, warning=FALSE}
cov2cor(vcov(m_heights))
```

Let's plot the posterior in 3D, because we **can**:

```{r posterior-3d-correct, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)
library(MASS)  # For kernel density estimation

# Extract posterior samples
posterior_samples <- extract.samples(m_heights)

# Calculate bivariate density using kde2d
density_est <- kde2d(posterior_samples$mu, posterior_samples$sigma, n = 100)

# Create a 3D surface plot
plot_ly(
  x = density_est$x,
  y = density_est$y,
  z = density_est$z,
  type = "surface",
  colorscale = "Viridis"
) %>%
  layout(
    title = "3D Posterior Density of μ and σ",
    scene = list(
      xaxis = list(title = "μ (mu)"),
      yaxis = list(title = "σ (sigma)"),
      zaxis = list(title = "Density")
    )
  )
```

**How beautiful ist that?**

We see in the 3D plot, that the "mountain" is not rotated, indicating 
graphically that the parameters are independent from each other.

We also see in the correlation matrix, the correlation of the parameters is $\sim 0$.
In the [context of a joint normal distribution](https://en.wikipedia.org/wiki/Correlation#Correlation_and_independence), 
this means that the parameters are independent. 

And, it is not an accident that the posterior looks like this. Using quadratic approximation,
we used the bivariate normal distribution to **approximate** the posterior.

## Classical approach for the simplest model

We have seen, how we could use domain and prior knowledge to fit a very simple model
for body heights of a population (!Kung San) in the Bayesian framework.

Now, let's start at the same point in the classical framework.
Here, we do not use any prior knowledge, at least not that explicitely.

The classical approach to fit a regression line is the so-called 
**[least squares method](https://en.wikipedia.org/wiki/Least_squares)**.

The **(simple mean-) model** is:

$$ Y_i = height_i = c + \varepsilon_i $$

- for some $c \in \mathbb{R}$ and
- normally distributed errors $\varepsilon_i \sim \text{Normal}(0, \sigma)$.

These are on average zero and have a constant standard deviation of $\sigma$.
So, we assume there is a fixed, but unknown, constant $c$ that we want to estimate and we 
assume that there is a special sort of error in our model that is normally distributed.
Sometimes there is a large deviation from the true $c$, sometimes there is a small deviation.
On average, the deviations are zero and the **errors should also be independent from each other**:
$$ \varepsilon_i \perp \varepsilon_j \text{ for } i \neq j$$
This means that just because I have just observed a large deviation from the true $c$
does not mean, that the probability of a large deviation in the next observation is higher/lower.
Note, that we cannot readily define different types of errors in the classical framework.

But what is $c$? We determine the shape of the model 
ourselves (constant model, or mean model) and then estimate the parameter $c$.
By defining the shape of the model ourselves and imposing a distribution where we want to 
estimate the parameter of said distribution, we are in **parametric statistics**.

We choose the $c$ which minimizes the sum of squared errors from the actual heights.
This has the advantage that deviations upper and lower from the actual height are
equally weighted. The larger the deviation the (quadratically) larger the penalty.

**Why do we do that?** Because, if the model assumptions (more on that later) 
are correct, the least squares
estimator is a really good estimator. How good? Later...

We want to miminize the following function:

$$ SSE \text{ (Sum of Squared Errors) }(c) = (height_1 - c)^2 + (height_2 - c)^2 + 
\ldots + (height_n - c)^2 =$$
$$ = \sum_{i=1}^{n} (height_i - c)^2$$

The SSE is a function of $c$ and we want to find the $c$ that minimizes the function.
Since it is a quadratic function, we can always find the minimum.
We have learnt in school how to do this (hopefully): Take the derivative of the function
and set it to zero. Solve for $c$ and you have the $c$ which yields the minimum of SSE(c).

Let's do that:

$$ \frac{d}{dc} SSE(c) =  2(height_1 - c)(-1) + 2(height_2 - c)(-1) + 
\ldots + 2(height_n - c)(-1) =$$
$$ = -2 \sum_{i=1}^{n} (height_i - c)$$

This should be zero for the minimum:
$$ -2 \sum_{i=1}^{n} (height_i - c) = 0$$
$$ \sum_{i=1}^{n} (height_i - c) = 0$$
$$ \sum_{i=1}^{n} (height_i - c) = 0$$
$$ \sum_{i=1}^{n} height_i - n \cdot c = 0$$
$$ \hat{c} = \frac{1}{n} \sum_{i=1}^{n} height_i = \overline{height_i}$$

The hat over the $c$ indicates that this is the estimated value of $c$.
Everytime we estimate a parameter, we put a hat over it.

And voilà, we have estimated the parameter $c$ of the model, which is just the 
sample mean of all the heights. In contrast to before, we did not put 
in a lot of prior knowledge, but just estimated the parameter from the data.

In R, we can do this easily:
```{r, echo=TRUE, , message=FALSE, warning=FALSE}
mod <- lm(height ~ 1, data = d2)
summary(mod)
dim(d2)
mean(d2$height) # same as the intercept
sd(d2$height) / sqrt(nrow(d2)) # standard error of the estimator
# test-statistic for the intercept:
mean(d2$height) / (sd(d2$height) / sqrt(nrow(d2)))
# residual standard error:
sqrt(sum(mod$residuals^2) / (nrow(d2) - 1))
```

the `~1` means that there is just a so-called **intercept** in the model.
There are **no covariates**, just the constant $c$.
This is the simplest we can do. `lm` stands for linear model and with this base 
command in R we ask the software to do the least squares estimation for us.

Let's look at the **R-output** of the model estimation:

- `lm(formula = height ~ 1, data = d2)`: This is the model we estimated.
- `Residuals`: The difference between the actual height and the estimated height: 
$r_i = height_i - \hat{c}$. A univariate 5-point summary is given.
- `Coefficients`: The estimated coefficients of the model. In this case, there is just the intercept.
   We get the 
  - `Std. Error` of the estimate, i.e. the [standard error](https://en.wikipedia.org/wiki/Standard_error) of the mean, 
    which is (according to the Central Limit Theorem) 
    $$\frac{\sigma}{\sqrt{n}}$$ and can be estimated by the sample 
    standard deviation divided by the square root of the sample size.
  - the `t value` and the `Pr(>|t|)` which is the $p$-value of the (Wald-)test of 
    the null hypothesis that the 
    coefficient is zero ($H_0: \text{intercept}=0$). 
    This is a perfect example of an absolutely useless $t$-test. 
    Why? Because obviously ([exercise 2](#exercise2_Intro)) the population mean of body heights is not zero.
- `Residual standard error`: The standard deviation of the residuals $r_i = height_i - \hat{c}$.
    In this case identical with the sample standard deviation 
    of heights ([exercise 3](#exercise3_Intro)). $351$ degrees of freedom. There are $352$ observations and 
    $1$ parameter estimated (intercept/mean). Hence, there are $352-1=351$ 
    freely movable variables in the statistic of the sample standard deviation.

Let's look at the situation graphically:

```{r, echo=FALSE, , message=FALSE, warning=FALSE}
library(rethinking)
library(tidyverse)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18, ]

# Select a few points to show vertical deviations
selected_points <- c(10, 50, 100)

mean_height <- mean(d2$height)

data.frame(height_i = 1:nrow(d2), height = d2$height) %>% # nolint: seq_linter.
  ggplot(aes(x = height_i, y = height)) +
  geom_point() +
  geom_hline(yintercept = mean_height, color = "red") +
  annotate("text", x = max(1:nrow(d2)) + 10, # nolint: seq_linter.
           y = mean_height,
           label = expression(hat(height[i])),
           color = "red", hjust = 0.6, vjust = 1.5) +
  geom_segment(data = data.frame(
    height_i = selected_points,
    height = d2$height[selected_points],
    mean_height = mean_height
  ),
  aes(x = height_i, xend = height_i,
      y = mean_height, yend = height),
  color = "red", linetype = "dotted") +
  ggtitle("Simple mean model estimation using Least Squares") +
  ylab("height estimated by model") +
  theme(plot.title = element_text(hjust = 0.5))

```

Above, the heights are plotted against the index of the observation.
The variability of heights around the regression line (constant in this case) seems to 
stay constant, which is a good sign. We will call this **homoscedasticity** later.
The dashed vertical red lines show two residuals, the difference between the actual height
and the estimated height. The model-estimated heights ($\widehat{heights_i}$) 
are all identical and nothing but the mean of all heights.

Peter Westfall explains in his excellent book a conditional distribution 
approach to regression. I highly recommend reading the first chapters.

What does this mean in this context?


## Exercises

### [E] Exercise 1 {#exercise1_Intro}

Use the [Swiss body heights](https://www.bfs.admin.ch/asset/de/30305714) data to 
determine 
- the 95% "Vertrauensintervall" for $\mu$ and
- calculate the standard deviation of the heights from $21,873$ Swiss people.

### [E] Exercise 2 {#exercise2_Intro}
Why do we not need a hypothesis test to know that the population mean of body heights is not zero?
Give 2 reasons.

### [M] Exercise 3 {#exercise3_Intro}
Verify analytically that the `Residual standard error` is identical with 
the sample standard deviation of the heights.

### [M] Exercise 4 {#exercise4_Intro}
Repeat the estimation of the simple model using a different data set about 
[chicken weights](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ChickWeight.html),
which is included in R.

- Set useful priors for the mean and standard deviation of the model 
  for the Baysian and the frequentist version considering 
  your a priori knowledge about chicken weights.

## Addendum

### The bivariate normal distribution {#bivariate_normal}

As a refresher, you can look into the old QM1 script and read the 
chapter "4.7 Gemeinsame Verteilungen". 
Maybe [this video](https://www.youtube.com/watch?v=SP2GKq8xJ5I&ab_channel=StatisticsNinja) 
also helps.

The bivariate normal distribution is a generalization of the normal distribution to two dimensions.
Now, we look at the distribution of two random variables $X$ and $Y$ **at the same time**.

Instead of one Gaussian bell curve, we have a 
[3D bell curve](https://en.wikipedia.org/wiki/Multivariate_normal_distribution#/media/File:Multivariate_Gaussian.png).
This curve defines how plausible different combinations of $X$ and $Y$ are.

Single points (like (3,6)) still have probability zero, because now the **volume** over a single point
($x$, $y$) is zero. The probability of a certain area is now the **volume** under the 
curve compared to the **area** under the density curve in the one-dimensional case.

**Example**: The following plot shows the density of a  bivariate normal distribution of 
two variables $X$ and $Y$
with $\mu_X = 0$, $\mu_Y = 0$, $\sigma_X = 1$, $\sigma_Y = 1$ and $\rho = \frac{2}{3}$.

Below is the correlation matrix of the bivariate normal distribution.

```{r, echo=FALSE, , message=FALSE, warning=FALSE}
library(MASS)    # For mvrnorm
library(plotly)  # For 3D plotting

# Define the parameters of the bivariate normal distribution
mu <- c(0, 0)                       # Mean
sigma <- matrix(c(0.75, 0.5, 0.5, 0.75), ncol = 2) # Covariance matrix

cov2cor(sigma)  # Convert covariance matrix to correlation matrix

# Generate a grid of x and y values
x <- seq(-3, 3, length.out = 100)
y <- seq(-3, 3, length.out = 100)
grid <- expand.grid(x = x, y = y)

# Compute the density of the bivariate normal
z <- matrix(dmvnorm(grid, mean = mu, sigma = sigma),
            nrow = length(x), ncol = length(y))

# Define the square for highlighting the probability volume
highlight_x <- c(0.5, 2)
highlight_y <- c(0.5, 2)

# Highlighted volume: Set values outside the square to NA
z_highlight <- z
z_highlight[grid$x < highlight_x[1] | grid$x > highlight_x[2] |
              grid$y < highlight_y[1] | grid$y > highlight_y[2]] <- NA

# Create a 3D plot
fig <- plot_ly() %>%
  add_surface(x = x, y = y, z = z, showscale = FALSE, opacity = 0.6) %>%
  add_surface(x = x, y = y, z = z_highlight,
              colorscale = "Viridis", opacity = 0.9) %>%
  layout(scene = list(
    xaxis = list(title = "X"),
    yaxis = list(title = "Y"),
    zaxis = list(title = "Density")
  ))

fig
```

If you move the plot around with your mouse, you see that there is a positive 
correlation between $X$ and $Y$ ($\rho = \frac{2}{3}$). This means that if $X$ is
above its mean, $Y$ is also more likely to be above its mean.
The variances of $X$ and $Y$ are both $1$. That means, that if you cut
through the plot in $X=0$ or $Y=0$, you see the same form of normal distribution.
If you look at if from above, we have hihglighted the section on the surface
over the area $X \in [0.5, 2]$ and $Y \in [0.5, 2]$. The volume over this area
under the density curve is the probability of this area: $P(X \in [0.5, 2] \text{ and } Y \in [0.5, 2])$
